---
title: "Metadata"
subtitle: "Best Practices in Environmental Data Science<br>FISH 549"
author: "Mark Scheuerell"
date: "6 February 2023"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align = "center", out.width = '90%')
```

## Goals for today

* Understand the concept of *metadata*

> - Understand the different kinds of metadata

> - Get an introduction to the *Ecological Metadata Language*


## What is metadata? | Definition

"data that provides information about other data"


## What is metadata? | Definition

"shorthand representation of the data to which they refer"


## Metadata is everywhere | Book

* title  
* author  
* publisher  
* copyright  
* table of contents / index  


## Metadata is everywhere | Phone calls

* number called or received from  
* time initiated & ended  
* location of callers  


## Metadata is everywhere | Email

* subject  
* from  
* to  
* date & time sent  
* sending & receiving IP addresses  


## Benefits of metadata

Metadata benefits science by

1) increased data longevity

    > - long-term studies outlast investigators/students
    
    > - many scientists contribute info from many areas


## Benefits of metadata

Metadata benefits science by

1) increased data longevity

2) increased data reuse & sharing 

    > - info helps others know if/how they could reuse them
    
    > - facilitates meta-analyses


## Benefits of metadata

Metadata benefits science by

1) increased data longevity

2) increased data reuse & sharing 

3) expanded scales/scopes of analyses

    > - short-term evolve into long-term
    
    > - facilitates creativity


## Kinds of metadata | Descriptive

*Descriptive metadata* is information about a resource that is used for searching & identification

> - title  
> - abstract  
> - author  
> - keywords


## Kinds of metadata | Structural

*Structural metadata* indicates how 2+ objects are connected

> - tables    
> - fields 
> - keys  
> - relationships  


## Entity–relationship model

Structural metadata are often conceptualized in an *entity–relationship model* 


## Entity–relationship model | Entity

An *entity* can be physical or logical

> - physical entities *actually* exist (eg, lake, fish)

> - logical entities *could* exist (eg, a sampling event)

> - entities have attributes (eg, depth, mass, date)


## Entity–relationship model | Relationship

Entities are connected via a *relationship*

> - relationships can be thought of as verbs

> - Sarah *supervises* Mark

> - Omar *samples* Lake Washington


## Entity–relationship diagram

ER models are usually drawn as boxes (entities) connected by lines (relationships)


## Entity–relationship diagram{data-background=images/ER_model_1.png data-background-size=90%}


## Entity–relationship diagram{data-background=images/ER_model_2.png data-background-size=90%}


## Kinds of metadata | Administrative

Administrative metadata refers to technical info about a data file

> - file type (eg, `.csv`)  
> - when the file was created
> - how the file was created


## Administrative metadata | Rights management

> - Who has had custody/ownership of the data?  
> - What intellectual property rights must be observed?


## Administrative metadata | Preservation

> - Where are the data stored?  
> - How are they identified (eg, digitial object identifier)?  


## Kinds of metadata | Reference

Information about the contents and quality of statistical data

> - description of fields  
> - QA/QC


## Kinds of metadata | Process

Describes the collection and any processing of the data

> - raw vs cleaned
> - reproducible workflows


## Process metadata | Michener (2005){data-background=images/eco_metadata.png data-background-position="50% 60%" data-background-size=90%}


## Kinds of metadata | Accessibility

Information about improved access to data

> - audio transcripts  
> - alternate text for an image  
> - large print


## Section 508{data-background=images/section508.png data-background-position="50% 60%" data-background-size=70%}


## How much metadata is enough?

Two factors to consider

1) effort involved in creating the metadata

2) value derived from it

In general, assume that "more is better"


## No metadata{data-background=images/metadata_ex1.png data-background-position="50% 40%" data-background-size=90%} 


## Minimal metadata{data-background=images/metadata_ex2.png data-background-position="50% 40%" data-background-size=90%} 


## Extensive{data-background=images/metadata_ex3.png data-background-position="70% 50%" data-background-size=50%} 


# Ecological Metadata Language


## Ecological Metadata Language | Fegraus et al (2005)

EML describes a range of essential aspects of ecological data

> - names & definitions of variables  
> - units of measurement  
> - date/time/location of data collection  
> - identity/contact for individual who collected the data  
> - sampling design


## Ecological Metadata Language | Fegraus et al (2005)

EML reduces ambiguity & uncertainty by formalizing metadata concepts

> - uses comprehensive **and** standardized set of terms  
> - uses definitions intended specifically for ecological data


## Ecological Metadata Language | Categories

1) General dataset

2) Geographic

3) Temporal

4) Taxonomic

5) Methods

6) Data table


## Ecological Metadata Language | General dataset

Contains concepts that

> - identify and name the dataset  
> - describe the purpose of the data collection  
> - describe the questions the data were intended to address


## General{data-background=images/metadata_general.png data-background-position="80% 50%" data-background-size=65%} 


## Ecological Metadata Language | Geographic

Contains information about *where*

> - the research project took place  
> - the samples were collected  
> - spatial or geographic references (UTM, lat/lon)


## Geographic{data-background=images/metadata_geo.png data-background-position="50% 40%" data-background-size=85%} 


## Ecological Metadata Language | Temporal

Contains information about *when*

> - range of dates (monthly between June 2019 and Dec 2020)  
> - specific time periods (01 May 2019, 08:00–12:00)  
> - gaps in time (no data from July 2020 because of power loss)


## Temporal{data-background=images/metadata_temporal.png data-background-position="50% 30%" data-background-size=85%} 


## Ecological Metadata Language | Taxonomic

Contains information about

> - taxonomic authority (book or system used to identify species)  
> - taxonomic rank (family, genus, species)


## Taxonomic{data-background=images/metadata_taxon.png data-background-position="50% 60%" data-background-size=85%} 


## Ecological Metadata Language | Methods

Contains information about *what happened*

> - instruments or devices used to collect data  
> - protocols
> - units of the samples

## Ecological Metadata Language | Methods

Contains information about *what happened*

* instruments or devices used to collect data  
* protocols
* units of the samples

Unlike the Methods section of a publication, *fully detailed descriptions* can be included


## Methods{data-background=images/metadata_methods.png data-background-position="50% 50%" data-background-size=85%} 


## Ecological Metadata Language | Data table

Contains information about a rectangular table

> - file name  
> - number of records
> - structure of the table (attributes of fields/columns)


## Ecological Metadata Language | Data table

> - *name* is a unique name for the field/column (`date`)  
> - *label* describes the field/column ("date of sample collection)  
> - *definition* indicates what the values represent (length of a fish)  
> - *units* (grams, meters)  
> - *type* (`numeric`, `character`)  
> - *precision* (mm)  
> - *attribute domain description* defines codes & domain of values
  > - `BVA = Bear Valley Creek`  
  > - `Length` is a positive, real value
  

## Data table{data-background=images/metadata_data.png data-background-position="50% 60%" data-background-size=70%} 


## Knowledge Network for Biocomplexity

The [KNB](https://knb.ecoinformatics.org/) is an international repository intended to facilitate ecological & environmental research

> - you can upload/publish data with a DOI

> - you can write EML


## &nbsp;{data-background=images/KNB.png data-background-position="50% 50%" data-background-size=100%} 


## [\{dataone\}](https://github.com/DataONEorg/rdataone) package

Provides read & write access to data and metadata from [DataONE network](https://www.dataone.org/)

> - KNB Data Repository  
> - Dryad  
> - NEON
> - USGS
> - NOAA


## [\{EML\}](https://docs.ropensci.org/EML/index.html) package

```{r eval = FALSE}
## username
me <- list(individualName = list(givenName = "Mark",
                                 surName = "Scheuerell"))

## list of attributes
my_eml <- list(
  dataset = list(
    title = "A Minimal Valid EML Dataset",
    creator = me,
    contact = me
  )
)

## inspect the EML
my_eml
```


## \{EML\} output {.smaller}

```
$dataset
$dataset$title
[1] "A Minimal Valid EML Dataset"

$dataset$creator
$dataset$creator$individualName
$dataset$creator$individualName$givenName
[1] "Mark"

$dataset$creator$individualName$surName
[1] "Scheuerell"



$dataset$contact
$dataset$contact$individualName
$dataset$contact$individualName$givenName
[1] "Mark"

$dataset$contact$individualName$surName
[1] "Scheuerell"
```


## What's next?

We'll discuss data analysis and visualization

