---
title: "Estimating species interactions"
subtitle: "SOE 592 – Intro to Time Series Analysis"
date: "<br>4 Jan 2024"
output:
  html_document:
    theme: simplex
    highlight: textmate
    css: ["lecture_inst.css", "fontawesome.css", "solid.css"]
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  fig.dim = c(6, 6),
  fig.align = "center"
  )
set.seed(592)
```

***

# Introduction

There are numerous ways to estimate the strength of ecological interactions among members of a food web, including top-down, bottom-up, and competitive interactions. These methods include experimental manipulation, stable isotope analysis, bioenergetics, and observational studies Back in the 1990s, a group of researchers began working on multivariate time series models that would allow people to estimate interactions among community members based on density (biomass) estimates collected over time. Among the best examples, is a 2003 paper by Tony Ives and colleagues, which you can find [here](https://doi.org/10.1890/0012-9615(2003)073[0301:ECSAEI]2.0.CO;2).


## State model for species interactions

Earlier we saw that we can use a discrete-time Gompertz model to estimate the intrinsic rate of increase and density-dependent effects on population growth. Here we extend the Gompertz model to a multispecies framework. For 2+ interacting species (taxa), we can write

$$
x_{i,t} = u_i + \sum^{m}_{j = 1}{b_{i,j} x_{j,t}} + w_t
$$

where $u_i$ is the intrinsic (density independent) growth rate, $b_{i,j}$ is the effect of the $j^{th}$ species on species $i$, and $b_{i,j}$ with $i = j$ is the density-dependent effect. We can write this model in matrix notation as

$$
\mathbf{x}_t = \mathbf{u} + \mathbf{B} \mathbf{x}_{t-1} + \mathbf{w}_t
$$

with

$$
\mathbf{u} =
\begin{bmatrix}
u_1 \\
u_2 \\
\vdots \\
u_m
\end{bmatrix}
~ ~ ~ ~ ~ ~ ~
\mathbf{B} =
\begin{bmatrix}
b_{1,1} & b_{1,2} & \dots & b_{1,m} \\
b_{2,1} & b_{1,2} & \dots & b_{2,m} \\
\vdots & \vdots & \ddots & \vdots \\
b_{m,1} & b_{m,2} & \dots & b_{m,m}
\end{bmatrix}
$$

## Covariate effects

We can easily accommodate the effects of exogenous drivers (covariates) by including two additional terms in the model, such that

$$
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{u} + \mathbf{C} \mathbf{c}_{t-k} + \mathbf{w}_t
$$

with the lag $k \geq 0$. The $m \times p$ matrix $\mathbf{C}$ contains the effect(s) of each covariate (cols) on each state (rows). The $p \times 1$ column vector $\mathbf{c}_{t-k}$ contains each of the $p$ covariates at time $t - k$.

The effect(s) of covariates can vary by state/process, species, etc, such that these are both possibilites (that could be compared via some information criterion like AICc):

$$
\mathbf{C} =
\begin{bmatrix}
C_{1, Temp} & C_{1, Rain} \\ 
C_{2, Temp} & C_{2, Rain} \\ 
\vdots & \vdots \\ 
C_{m, Temp} & C_{m, Rain}
\end{bmatrix}
~~ \text{or} ~~
\mathbf{C} =
\begin{bmatrix}
C_{Temp} & C_{Rain} \\ 
C_{Temp} & C_{Rain} \\ 
\vdots & \vdots \\ 
C_{Temp} & C_{Rain}
\end{bmatrix}
$$

with

$$
\mathbf{c}_{t-k} =
\begin{bmatrix}
Temp_{t-k} \\
Rain_{t-k}
\end{bmatrix}
$$


## A note on $\mathbf{u}$

Recall that when we discussed univariate state-space models, we saw that it's ***very*** difficult to estimate both $\mathbf{u}$ and $\mathbf{B}$

We will typically subtract the mean from each time series and assume $\mathbf{u}$ = 0


***

# Wolves & moose

Population dynamics of wolves and moose on Isle Royale, Michigan make an interesting case study of a two-species predator-prey interactions.  These populations have been studied intensively since 1958.  Unlike other populations of gray wolves, the Isle Royale population has a diet dominated by one prey item, moose. The only predator of moose on Isle Royale is the gray wolf, as this population is not hunted.  

We will use the wolf and moose winter census data from Isle Royale to learn how to fit community dynamics models to time series data.  The long-term January (wolf) and February (moose) population estimates are provided by the Isle Royale Wolf-Moose Project. You can find out more about the organization and the project [here](http://www.isleroyalewolf.org).

The mathematical form of the process model for the wolf-moose population dynamics is

$$
\begin{equation}
\begin{gathered}
\begin{bmatrix}
  x_{w}\\x_{m}
\end{bmatrix}_t =
\begin{bmatrix}
  b_{w \rightarrow w} & b_{m \rightarrow w} \\
  b_{w \rightarrow m} & b_{m \rightarrow m}
\end{bmatrix} 
\begin{bmatrix}
  x_{w} \\
  x_{m}
\end{bmatrix}_{t-1} + 
\begin{bmatrix}
  u_{w} \\
  u_{m}
\end{bmatrix} +
\begin{bmatrix}
  w_{w} \\
  w_{m}
\end{bmatrix}_t \\
\begin{bmatrix}
  w_{w} \\
  w_{m}
\end{bmatrix}_t 
\sim
\text{MVN}
\begin{pmatrix}
  0,
  \begin{bmatrix}
    q_{w} & 0 \\
    0 & q_{m}
  \end{bmatrix}
\end{pmatrix}
\end{gathered}
\end{equation}
$$

where $w$ denotes wolf and $m$ denotes moose. $w \rightarrow w$ is the effect of wolf on wolf (density-dependence) and $w \rightarrow m$ is the effect of wolf on moose (predation effect on moose).

## Get the data

The data are available in the {MARSS} package, so we won't have to download them the project site; we'll use the years 1960 to 2011.

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Get the data and plot the time series.
</div>

```{r get-data}
## load MARSS library
library(MARSS)

## set the years
yr_set <- seq(1960, 2011) #isleRoyal[, "Year"] >= 1960 & isleRoyal[, "Year"] <= 2011

## select the wolf & moose data & log-transform them
royale_data <- isleRoyal[isleRoyal[,"Year"] %in% yr_set,
                         c("Wolf", "Moose")] |>
  log()

## set of years
tt <- isleRoyal[, "Year"]

## plot the data
par(mai = c(1.2, 1, 0.3, 0),
    omi = c(0, 0, 0.5, 1))
matplot(yr_set, royale_data,
  ylab = "Log of counts", xlab = "Year", type = "o", pch = 16,
  lwd = 2, col = c("darkgray", "brown")
)
legend(x = "center", legend = c("Moose", "Wolves"),
       col = c("brown", "darkgray"), lwd = 2, bty = "n")
```


## Fit the model

Fitting our species interaction model with `MARSS()` is similar to how we've done it for other problems. The key is to set up the model list correctly and everything should work. Thus, we need to explicitly specify the forms for ***all*** of the vectors and matrices in the full MARSS model given by

$$
\mathbf{x}_t = \mathbf{B} \mathbf{x}_{t-1} + \mathbf{u} + \mathbf{C} \mathbf{c}_{t-k} + \mathbf{w}_t \\
\mathbf{y}_t = \mathbf{Z} \mathbf{x}_t + \mathbf{a} + \mathbf{D} \mathbf{d}_{t-h}+ \mathbf{v}_t \\
$$

Thus, to fit our species interaction model, we need to set the following:

$$
\begin{align}
\mathbf{C} &=
\begin{bmatrix}
0 \\ 0
\end{bmatrix} \\
\mathbf{c}_{t-k} &=
\begin{bmatrix}
0
\end{bmatrix} \\
\mathbf{Z} &= 
\begin{bmatrix}
1 & 0 \\ 0 & 1
\end{bmatrix} \\
\mathbf{a} &=
\begin{bmatrix}
0 \\ 0
\end{bmatrix} \\
\mathbf{D} &=
\begin{bmatrix}
0 \\ 0
\end{bmatrix} \\
\mathbf{d}_{t-h} &=
\begin{bmatrix}
0
\end{bmatrix}
\end{align}
$$

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Set up the model list for `MARSS()` (the code below uses textual shortcuts).
</div>

```{r define-model-1}
## create model list for MARSS()
mod_list <- list(
  ## state model
  B = "unconstrained",
  U = "unequal",
  C = "zero",
  c = "zero",
  Q = "diagonal and unequal",
  ## obs model
  Z = "identity",
  A = "zero",
  D = "zero",
  d = "zero",
  R = "diagonal and unequal"
  )
```

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Fit the species interaction model define above with `MARSS()`.
</div>

```{r fit-model-1, cache = TRUE}
## transform the data for MARSS()
royale_data <- t(royale_data)

## fit the model
mod_first <- MARSS(royale_data, model = mod_list)
```

<div class="boxy boxy-red boxy-exclamation">
**Note:** This model has not converged because we're trying to estimate both $\mathbf{B}$ and $\mathbf{u}$, which are highly confounded.
</div>

<div class="boxy boxy-orange boxy-lightbulb">
**Tip:** We can assume the populations are stationay about the mean and subtract the mean from each time series. We can then assume $\mathbf{u}$ = 0.
</div>

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Use the `zscore()` function from the {MARSS} package to remove the mean and also standardize the variance.
</div>

```{r z-score-data}
## z-score the data
royale_data_z <- zscore(royale_data)
```

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Update the model list for `MARSS()` by setting $\mathbf{u}$ to **0** and then refit the model.
</div>

```{r fit-model-2, cache = TRUE}
## set u vector to all 0's
mod_list$U <- "zero"

## refit model
mod_second <- MARSS(royale_data_z, model = mod_list) #, control = cntl.list)
```

<div class="boxy boxy-red boxy-exclamation">
**Note:** This model is still not converging, but this time it looks like the terms in $\mathbf{R}$ for the observation variance are heading toward 0.
</div>

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Go back and inspect the plot of the log-counts for moose and wolves. Does it look smooth like a random walk or choppy like white noise?
</div>

We could try increasing the minimum number iterations via the `control` argument to `MARSS()` like we've done before, but that probably won't help in this case. Instead, let's assume the animals on the island are relatively easy to count and that the data represent a complete census.

<div class="boxy boxy-blue boxy-clipboard-list">
**Task:** Set $\mathbf{R}$ equal to **0** in the model list and refit the model.
</div>

```{r fit-model-3, cache = TRUE}
## set the observation variance to 0
mod_list$R <- "zero"

## refit the model
mod_third <- MARSS(royale_data_z, model = mod_list)
```

## Examine the interaction effects

The estimated $\mathbf{B}$ elements 

```{r model-coef}
B_hat <- coef(mod_third, type = "matrix")$B
rownames(B_hat) <- colnames(B_hat) <- rownames(royale_data)
print(B_hat, digits = 2)
```

The function returns the estimated parameters, but in this case we want to see the estimates in matrix form. Element row=$i$, col=$j$ in $\mathbf{B}$ is the effect of species $j$ on species $i$, so $\mathbf{B}_{2,1}$ is the effect of wolves on moose and $\mathbf{B}_{1,2}$ is the effect of moose on wolves. The $\mathbf{B}$ matrix suggests that wolves have a negative effect on moose and that moose have a positive effect on wolves---as one would expect.  The diagonals are interpreted differently than the off-diagonals since the diagonals are ($b_{i,i}-1$) so subtract off 1 from the diagonals to get the effect of species $i$ on itself.   If the species are density-independent, then $\mathbf{B}_{i,i}$ would equal 1.  Smaller $\mathbf{B}_{i,i}$ means more density dependence.   

## Adding covariates

It is well-known that moose numbers are strongly affected by winter and summer climate.  The Isle Royale data set provided with MARSS has climate data from climate stations in Northeastern Minnesota, near Isle Royale.  The covariate data include January-February, July-September and April-May average temperature and precipitation.  Also included are three-year running means of these data, where the number for year $t$ is the average of years $t-1$, $t$ and $t+1$.  We will include these covariates in the analysis to see how they change our interaction estimates.  We have to adjust our covariates because the census numbers are from winter in year $t$ and we want the climate data from the previous year to affect this winter's moose count.  As usual, we will need to demean our covariate data so that we can set $\mathbf{u}$ equal to zero.  We will standardize the variance also so that we can more easily compare the effects across different covariates.

The mathematical form of our new process model for the wolf-moose population dynamics is

$$
\begin{equation}
\begin{gathered}
\begin{bmatrix}
  x_{w} \\
  x_{m}
\end{bmatrix}_t =
\mathbf{B} 
\begin{bmatrix}
  x_{w} \\
  x_{m}
\end{bmatrix}_{t-1} + 
\begin{bmatrix}
  0 & 0 & 0 \\
  C_{21} & C_{22} & C_{23}
\end{bmatrix} 
\begin{bmatrix}
  \text{win temp} \\
  \text{win precip} \\
  \text{sum temp}
\end{bmatrix}_{t-1} +
\begin{bmatrix}
  w_{w} \\
  w_{m}
\end{bmatrix}_t
\end{gathered}
\end{equation}
$$

The $C_{21}$, $C_{22}$, etc. terms are the effect of winter temperature, winter precipitation, previous summer temperature and previous summer precipitation on winter moose numbers.  Since climate is known to mainly affect the moose, we set the climate effects to 0 for wolves (top row of $\mathbf{C}$).  

First we prepare the covariate data and select the winter temperature and precipitation data and the summer temperature  data.  We need to use the previous year's climate data with this winter's abundance data, so 1959 to 2010.

```{r}
clim.variables <- c(
  "jan.feb.ave.temp", "jan.feb.ave.precip",
  "july.sept.ave.temp"
)
yr1959to2010 <- isleRoyal[, "Year"] >= 1959 & isleRoyal[, "Year"] <= 2010
clim.dat <- t(isleRoyal[yr1959to2010, clim.variables])
z.score.clim.dat <- zscore(clim.dat)
```

A plot of the covariate data against each other indicates that there is not much correlation between winter temperature and precipitation, which is good for analysis purposes, but warm winters are somewhat correlated with warm summers.  The latter will make it harder to interpret the effect of winter versus summer temperature although the correlation is not too strong fortunately.

Next we prepare the list with the structure of all the model matrices.  We give descriptive names to the $\mathbf{C}$ elements so we can remember what each $\mathbf{C}$ element means.

```{r}
royale.model.3 <- list(
  Z = "identity", B = "unconstrained",
  Q = "diagonal and unequal", R = "zero", U = "zero",
  C = matrix(list(
    0, "Moose win temp", 0, "Moose win precip",
    0, "Moose sum temp"
  ), 2, 3),
  c = z.score.clim.dat
)
```

Then we fit the model with covariates.

```{r}
kem.3 <- MARSS(royale_data_z, model = royale.model.3)
```

The results suggest what is already known about this system: cold winters and heavy snow are bad for moose as are hot summers.


```{r}
cor.fun <- function(x, y) {
  text(0.5, 0.5, format(cor(x, y), digits = 2), cex = 2)
}
pairs(t(z.score.clim.dat), lower.panel = cor.fun)
```


## Change the model and data

You can explore the sensitivity of the $\mathbf{B}$ estimates when the measurement error is increased by adding white noise to the data:

```{r, eval = FALSE}
bad.data <- royale_data_z + matrix(rnorm(100, 0, sqrt(.2)), 2, 50)
kem.bad <- MARSS(bad.data, model = model)
```

You can change the model by changing the constraints on $\mathbf{R}$ and $\mathbf{Q}$.

## Improving performance

In the default MARSS model, the value of $\mathbb{E}[\mathbf{x}_0|\mathbf{y}_0]$ (what `x0` denotes when `tinitx = 0` in the model list) is estimated.  If we are estimating the $\mathbf{B}$ matrix, it is better to set `tinitx = 1` so that we are estimating $\mathbb{E}[\mathbf{x}_1|\mathbf{y}_0]$ instead.  The model will fit either way, but setting `tinitx = 1` in the model list will speed up and stabilize the fitting. It does not make much of a difference for the wolf-moose dataset but can have a large effect for larger models.  The reason is that the likelihood surface for $\mathbb{E}[\mathbf{x}_1|\mathbf{y}_0]$ is better behaved when $\mathbf{B}$ is small.  For example, if $\mathbf{B}$ equal to 0, there is little information about $\mathbb{E}[\mathbf{x}_0|\mathbf{y}_0]$ so the algorithm goes in circles trying to estimate it while there is good information $\mathbb{E}[\mathbf{x}_1|\mathbf{y}_0]$ from $\mathbf{y}_1$.  We could use a prior on the initial $\mathbf{x}$ but this requires its variance-covariance structure, which depends on the unknown $\mathbf{B}$ and specifying a variance-covariance structure that conflicts with $\mathbf{B}$ will change your $\mathbf{B}$ estimates.\index{troubleshooting!sensitivity to x0 prior}

For the wolf-moose model, we set $\mathbf{R}=0$.  The EM algorithm (default) cannot estimate `x0` when `tinitx = 1` therefore to fit the model with `tinitx = 1` we need to use `method = "BFGS"`.  This is only for the case when $\mathbf{R} = 0$.

```{r}
royale.model.4 <- list(
  B = "unconstrained", U = "zero", Q = "diagonal and unequal",
  Z = "identity", R = "zero", tinitx = 1
)
kem.4 <- MARSS(royale_data_z, model = royale.model.4)
```

\index{troubleshooting!degenerate variances}The other setting we may want to change is `allow.degen` in the control list.  This sets the diagonals of $\mathbf{Q}$ or $\mathbf{R}$ to zero if they are heading towards zero.  When the initial $\mathbf{x}$ is at $t = 1$, this can have non-intuitive (not wrong but puzzling) consequences if $\mathbf{R}$ is going to zero.   So, we will set `control$allow.degen = FALSE` and manually set $\mathbf{R}$ to $\mathbf{0}$ if needed.
