---
title: "Analyzing animal movement"
subtitle: "SOE 592 – Intro to Time Series Analysis"
date: "<br>4 Jan 2024"
output:
  html_document:
    theme: simplex
    highlight: textmate
    css: ["lecture_inst.css", "fontawesome.css", "solid.css"]
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE, purl=FALSE}
## in case you forget to add a chunk label
knitr::opts_chunk$set(
  fig.dim = c(8, 5),
  fig.align = "center"
  )
set.seed(592)
```

***

# Introduction

Using time series models to analyze animal movement is common in wildlife ecology. The topic has seen increasingly more interest as the technology for tracking animals has become more sophisticated and less expensive. In this lab, we'll see how to use multivariate state-space models to estimate movement tracks and rates of travel for several different behaviors.


***

# Wandering

Sometimes animals appear to be merely wandering around in search of food, mates, or resting spots. For these types of behavior, random walks models are most appropriate. For terrestrial animals, most movement models consider locations in two dimensions, latitude and longitude, and generally ignore the height above the ground. For some fishes and marine mammals, however, the models may additionally include the animal's depth below the surface of the water. Here we will consider only two-dimensional movement, but the concepts are the same when applied to three dimensions.

## Random walk model

Recall that random walk models are extremely flexible in their ability to fit a variety of temporal patterns. As such, random walk models work really well for animals that appear to be wandering around with a lack of clear direction or a tendency to stay around some particular area.

### State model

We can write the state model for movement in two dimensions as

$$
x^\text{(lat)}_t = x^\text{(lat)}_{t-1} + w^\text{(lat)}_t \\
x^\text{(lon)}_t = x^\text{(lon)}_{t-1} + w^\text{(lon)}_t \\
\Downarrow \\
\begin{bmatrix}
x^\text{(lat)} \\
x^\text{(lon)} \\
\end{bmatrix}_t =
\begin{bmatrix}
x^\text{(lat)} \\
x^\text{(lon)} \\
\end{bmatrix}_{t-1} + 
\begin{bmatrix}
w^\text{(lat)} \\
w^\text{(lon)} \\
\end{bmatrix}_t \\
\Downarrow \\
\mathbf{x}_t = \mathbf{x}_{t-1} + \mathbf{w}_t \\
$$

where the process errors are distributed as $\mathbf{w}_t \sim \text{MVN}(\mathbf{0}, \mathbf{Q})$.

<div class="boxy boxy-orange boxy-lightbulb">
**Tip:** The form for $\mathbf{Q}$ can be changed to account for more/less variance in the latitude or longitude step size (ie, `diagonal and unequal` in `MARSS()`), or to model covariance between the errors (ie, `equalvarcov` in `MARSS()`).
</div>

### Observation model

Our observation model is straightforward and includes terms for the data, states, and errors.

$$
y^\text{(lat)}_t = x^\text{(lat)}_t + v^\text{(lat)}_t \\
y^\text{(lon)}_t = x^\text{(lon)}_t + v^\text{(lon)}_t \\
\Downarrow \\
\begin{bmatrix}
y^\text{(lat)} \\
y^\text{(lon)} \\
\end{bmatrix}_t =
\begin{bmatrix}
x^\text{(lat)} \\
x^\text{(lon)} \\
\end{bmatrix}_t + 
\begin{bmatrix}
v^\text{(lat)} \\
v^\text{(lon)} \\
\end{bmatrix}_t \\
\Downarrow \\
\mathbf{y}_t = \mathbf{x}_t + \mathbf{v}_t \\
$$

where the process errors are distributed as $\mathbf{v}_t \sim \text{MVN}(\mathbf{0}, \mathbf{R})$.

<div class="boxy boxy-orange boxy-lightbulb">
**Tip:** It's rare that the variance in observation errors would be different for latitude or longitude so we typically assume that they're "independent and identically distributed" (IID), such that the form for $\mathbf{R}$ is `diagonal and equal` in `MARSS()`.
</div>



***

# Migration

## Biased random walk

$$
\begin{bmatrix}
x^\text{(lat)}_{t} \\
x^\text{(lon)}_{t} \\
\end{bmatrix} =
\begin{bmatrix}
x^\text{(lat)}_{t-1} \\
x^\text{(lon)}_{t-1} \\
\end{bmatrix} + 
\begin{bmatrix}
u^\text{(lat)} \\
u^\text{(lon)} \\
\end{bmatrix} + 
\begin{bmatrix}
w^\text{(lat)}_{t} \\
w^\text{(lon)}_{t} \\
\end{bmatrix}
$$


***

# Home range



# Model comparisons

## Evidence for bias

Q: Is there data support for the bias term?

To answer this, we can fit a model without the bias term and compare AICc


## Evidence for bias

We need to modify the model definition for `MARSS()` by setting `U = matrix(0)` 

```{r rw_model, echo=TRUE}
mod_list <- list(
  ## state model
  B = matrix(1), U = matrix(0), Q = matrix("q"),
  ## obs model
  Z = matrix(1), A = matrix(0), R = matrix("r")
  )
```


## Re-fit the model with `MARSS()`

```{r rw_fit, echo = TRUE, eval=TRUE}
## fit the model
# mod_fit_2 <- MARSS(y = YY, model = mod_list)
```


## Compare AIC values

```{r compare_AICc, echo = TRUE, eval = TRUE}
## biased RW
# mod_fit$AICc

## unbiased RW
# mod_fit_2$AICc
```

***

## A note on model fitting

When we have 2+ observations of process, our estimates of model parameters are much more accurate and precise.

$$
\begin{align}
x_t &= x_{t-1} + u + w_t ~~ \text{with} ~ w_t \sim \text{N}(0,q) \\
~ \\
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}_t
&=
\begin{bmatrix}
1 \\
1 \\
\vdots \\
1
\end{bmatrix}
x_t + 
\begin{bmatrix}
v_1 \\
v_2 \\
\vdots \\
v_n
\end{bmatrix}_t
~~ \text{with} ~
\mathbf{v_t} \sim \text{MVN}(\mathbf{0},\mathbf{R})
\end{align}
$$



