---
title: "Fitting univariate state-space models"
subtitle: "SOE 592 – Intro to Time Series Analysis"
date: "<br>4 Jan 2024"
output:
  html_document:
    theme: sandstone
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(
  fig.dim = c(7, 5),
  fig.align = "center"
  )
set.seed(592)
```

# Introduction


# Fitting models


## Simulate a Gompertz model

Simulate the AR(1) state model

```{r gompertz_sim, echo = TRUE, eval= TRUE}
## number of time steps
TT <- 40
## strength of density-dependence (0 < b < 1)
bb <- 0.5
## time series of process errors with SD = 1
ww <- rnorm(TT, 0, 0.5)
## initialize state & set x0 = w0
xx <- ww
## loop over time steps
for(t in 2:TT) {
  xx[t] <- bb * xx[t-1] + ww[t]
}
```


## Simulate a Gompertz model

Add some observation error

```{r gomp_obs_error, echo = TRUE, eval= TRUE}
## obs errors with SD = 0.5
vv <- rnorm(TT, 0, 0.5)
## obs data
yy <- xx + vv
```


## Fitting models with `MARSS()`

We can use the __{MARSS}__ pkg for R to fit state-space models

To do so, we must write the model out just as we would on a white board


## Fitting models with `MARSS()`

Here are the main function arguments

```{r, echo=TRUE, eval=FALSE}
MARSS(y, model = NULL, inits = NULL, control = NULL, ...)
```

* `y` is an $n \times T$ `matrix` of data (observations)

* `model` a `list` that defines the state-space model in terms of parameters to be estimated

* `inits` [_optional_] a `list` of initial values for parameters to be estimated

* `control` [_optional_] a `list` of options for controlling fitting algorithms, setting tolerance & convergence parameters, etc


## Fitting models with `MARSS()`

`MARSS()` uses an expanded form of state-space model, such that our biased random walk

$$
\begin{align}
x_t &= x_{t-1} + u + w_t \\
y_t &= x_t + v_t
\end{align}
$$
instead becomes

$$
\begin{align}
x_t &= b x_{t-1} + u + w_t \\
y_t &= Z x_t + a + v_t
\end{align}
$$

with $b = 1$, $Z = 1$, and $a = 0$


## Fitting models with `MARSS()`

State model  
$x_t = b x_{t-1} + u + w_t ~~ \text{with} ~~ w_t \sim \text{N}(0,q)$

Observation model  
$y_t = Z x_t + a + v_t ~~ \text{with} ~~ v_t \sim \text{N}(0,r)$

```{r, echo=TRUE}
mod_list <- list(
  ## state model
  B = matrix(1), U = matrix("u"), Q = matrix("q"),
  ## obs model
  Z = matrix(1), A = matrix(0), R = matrix("r")
  )
```

**Note**: `MARSS()` works with the `matrix` class in **R**, so we have to define any scalar as a $1 \times 1$ matrix


## Fitting models with `MARSS()`

What about the initial state $x_0$?

$$
x_0 \sim \text{N}(\pi,\lambda)
$$

Estimating both $\pi$ and $\lambda$ is nearly impossible, so the default is to treat $x_0$ as fixed, but unknown, effect rather than as a random effect, such that

$$
x_0 \sim \text{N}(\pi,0)
$$

`MARSS()` will do this for us


## Fitting models with `MARSS()`

We can ignore the `inits` and `control` arguments for now and fit the model

```{r, echo=TRUE, eval=FALSE}
## load MARSS package
library(MARSS)
## define the data as an N (rows) x T (cols) matrix
dat <- matrix(yy, nrow = 1, ncol = TT)
## fit the model
mod_fit <- MARSS(y = dat, model = mod_list)
```


## Fitting models with `MARSS()` {.smaller}

```{r brw_fit, echo=FALSE, eval=TRUE}
## load MARSS package
library(MARSS)
## define the data as an N (rows) x T (cols) matrix
dat <- matrix(yy, nrow = 1, ncol = TT)
## fit the model
mod_fit <- MARSS(y = dat, model = mod_list)
```


## A note on model fitting

When we have 2+ observations of process, our estimates of model parameters are much more accurate and precise.

$$
\begin{align}
x_t &= x_{t-1} + u + w_t ~~ \text{with} ~ w_t \sim \text{N}(0,q) \\
~ \\
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}_t
&=
\begin{bmatrix}
1 \\
1 \\
\vdots \\
1
\end{bmatrix}
x_t + 
\begin{bmatrix}
v_1 \\
v_2 \\
\vdots \\
v_n
\end{bmatrix}_t
~~ \text{with} ~
\mathbf{v_t} \sim \text{MVN}(\mathbf{0},\mathbf{R})
\end{align}
$$



## Extracting fitted model objects

Fitted values are stored in an $N \times T$ matrix (here $N = 1$) accessed with `$states`

Standard errors of fitted values are stored in an $N \times T$ matrix (vector) accessed with `$states.se`

```{r, echo = TRUE, eval = TRUE}
## T x 1 (transposed) vector of states
mod_fits <- t(mod_fit$states)

## T x 1 (transposed) vector of SE's
mod_fits_SE <- t(mod_fit$states.se)
```


## Approximate confidence intervals

We can estimate a $[(1 - \alpha)  \times 100]\%$ confidence interval as

$$
\hat{x} ~ \pm ~ t_{1 - \alpha/2} \times \text{SE}(\hat{x})
$$


## Approximate confidence intervals

```{r brw_CI, eval = TRUE, echo = TRUE}
## upper 95% CI
mod_fits_CI_hi <- mod_fits + qt(p = 0.975, df = TT - 1) * mod_fits_SE

## lower 95% CI
mod_fits_CI_lo <- mod_fits - qt(p = 0.975, df = TT - 1) * mod_fits_SE
```


<!-- ## Plot the model fit & uncertainty -->

<!-- Plot the state and observation -->

<!-- ```{r rw_plot_fit_code, echo=TRUE, eval=FALSE} -->
<!-- plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf", -->
<!--         las = 1, ylim = c(min(xx,yy), max(xx,yy)), -->
<!--         ylab = expression(italic(x[t])~~or~~italic(y[t]))) -->
<!-- lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870") -->
<!-- lines(mod_fits, lwd = 2, type = "l", cex = 1.5, col = "black") -->
<!-- lines(mod_fits_CI_hi, lwd = 2, type = "l", cex = 1.5, col = "gray") -->
<!-- lines(mod_fits_CI_lo, lwd = 2, type = "l", cex = 1.5, col = "gray") -->
<!-- ``` -->


## Model fit and true states 

```{r rw_plot_fits_states, align.fig="center"}
par(mai = c(0.8,0.8,0,0), omi = c(0,0,0,0))
plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylim = c(min(xx,yy), max(xx,yy)),
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
# lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870")
lines(mod_fits, lwd = 2, type = "l", cex = 1.5, col = "black")
lines(mod_fits_CI_hi, lwd = 2, type = "l", cex = 1.5, col = "gray")
lines(mod_fits_CI_lo, lwd = 2, type = "l", cex = 1.5, col = "gray")
```


## Model fit with observations only

```{r rw_plot_fit_nostate, align.fig="center"}
par(mai = c(0.8,0.8,0,0), omi = c(0,0,0,0))
plot.ts(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870",
        las = 1, ylim = c(min(xx,yy), max(xx,yy)),
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(mod_fits, lwd = 2, type = "l", cex = 1.5, col = "black")
lines(mod_fits_CI_hi, lwd = 2, type = "l", cex = 1.5, col = "gray")
lines(mod_fits_CI_lo, lwd = 2, type = "l", cex = 1.5, col = "gray")
```


## Model fit with observations & states 

```{r rw_plot_fit, align.fig="center"}
par(mai = c(0.8,0.8,0,0), omi = c(0,0,0,0))
plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylim = c(min(xx,yy), max(xx,yy)),
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870")
lines(mod_fits, lwd = 2, type = "l", cex = 1.5, col = "black")
lines(mod_fits_CI_hi, lwd = 2, type = "l", cex = 1.5, col = "gray")
lines(mod_fits_CI_lo, lwd = 2, type = "l", cex = 1.5, col = "gray")
```


## Evidence for bias

Q: Is there data support for the bias term?

To answer this, we can fit a model without the bias term and compare AICc


## Evidence for bias

We need to modify the model definition for `MARSS()` by setting `U = matrix(0)` 

```{r rw_model, echo=TRUE}
mod_list <- list(
  ## state model
  B = matrix(1), U = matrix(0), Q = matrix("q"),
  ## obs model
  Z = matrix(1), A = matrix(0), R = matrix("r")
  )
```


## Re-fit the model with `MARSS()`

```{r rw_fit, echo = TRUE, eval=TRUE}
## fit the model
mod_fit_2 <- MARSS(y = dat, model = mod_list)
```


## Compare AIC values

```{r compare_AICc, echo = TRUE, eval = TRUE}
## biased RW
mod_fit$AICc

## unbiased RW
mod_fit_2$AICc
```


## Fitting models with `MARSS()`

Recall that `MARSS()` uses an expanded form of state-space model, such that

$$
\begin{align}
x_t &= b x_{t-1} + w_t \\
y_t &= x_t + v_t \\
& \Downarrow \\
x_t &= b x_{t-1} + u + w_t \\
y_t &= Z x_t + a + v_t
\end{align}
$$

with $u = 0$, $Z = 1$, and $a = 0$


## Fitting models with `MARSS()`

State model  
$x_t = b x_{t-1} + u + w_t ~~ \text{with} ~~ w_t \sim \text{N}(0,q)$

Observation model  
$y_t = Z x_t + a + v_t ~~ \text{with} ~~ v_t \sim \text{N}(0,r)$

```{r, echo=TRUE}
mod_list <- list(
  ## state model
  B = matrix("b"), U = matrix(0), Q = matrix("q"),
  ## obs model
  Z = matrix(1), A = matrix(0), R = matrix("r")
  )
```

**Note**: `MARSS()` works with the `matrix` class in **R**, so we have to define any scalar as a $1 \times 1$ matrix


## Fitting models with `MARSS()`

We can ignore the `inits` and `control` arguments for now and fit the model

```{r, echo = TRUE, eval = FALSE}
## define the data as an N (rows) x T (cols) matrix
dat <- matrix(yy, nrow = 1, ncol = TT)
## fit the model
mod_fit <- MARSS(y = dat, model = mod_list)
```


## Fitting models with `MARSS()`

```{r gomp_fit, echo = FALSE, eval = TRUE}
## define the data as an N (rows) x T (cols) matrix
dat <- matrix(yy, nrow = 1, ncol = TT)
## fit the model 
mod_fit <- MARSS(y = dat, model = mod_list)
```


