---
title: "Analyzing animal movement"
subtitle: "SOE 592 – Intro to Time Series Analysis"
date: "<br>4 Jan 2024"
output:
  html_document:
    theme: sandstone
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r dfa-setup, include=FALSE, purl=FALSE}
## in case you forget to add a chunk label
knitr::opts_knit$set(unnamed.chunk.label = "dfa-")
```

# Introduction


***

# Model comparisons

## Evidence for bias

Q: Is there data support for the bias term?

To answer this, we can fit a model without the bias term and compare AICc


## Evidence for bias

We need to modify the model definition for `MARSS()` by setting `U = matrix(0)` 

```{r rw_model, echo=TRUE}
mod_list <- list(
  ## state model
  B = matrix(1), U = matrix(0), Q = matrix("q"),
  ## obs model
  Z = matrix(1), A = matrix(0), R = matrix("r")
  )
```


## Re-fit the model with `MARSS()`

```{r rw_fit, echo = TRUE, eval=TRUE}
## fit the model
mod_fit_2 <- MARSS(y = YY, model = mod_list)
```


## Compare AIC values

```{r compare_AICc, echo = TRUE, eval = TRUE}
## biased RW
mod_fit$AICc

## unbiased RW
mod_fit_2$AICc
```

***

## A note on model fitting

When we have 2+ observations of process, our estimates of model parameters are much more accurate and precise.

$$
\begin{align}
x_t &= x_{t-1} + u + w_t ~~ \text{with} ~ w_t \sim \text{N}(0,q) \\
~ \\
\begin{bmatrix}
y_1 \\
y_2 \\
\vdots \\
y_n
\end{bmatrix}_t
&=
\begin{bmatrix}
1 \\
1 \\
\vdots \\
1
\end{bmatrix}
x_t + 
\begin{bmatrix}
v_1 \\
v_2 \\
\vdots \\
v_n
\end{bmatrix}_t
~~ \text{with} ~
\mathbf{v_t} \sim \text{MVN}(\mathbf{0},\mathbf{R})
\end{align}
$$



