---
title: "Univariate<br>state-space models"
subtitle: "SOE 592 – Intro to Time Series Analysis"
author: "Mark Scheuerell"
date: "4 Jan 2024"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
set.seed(592)
```

## Topics

Univariate state-space models

1. state (process) model

2. observation model

Simple models of population dynamics


## State-space models

Consist of 2 parts


## Part 1: State model

Describes the __true state of nature__ over time

```{r state_model, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center'}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## boundaries
ss <- 5
nn <- 7
rr <- ss*3
cc <- ss*nn
## mid-points
xm <- ss/2 + seq(0,cc-ss,ss)
ymt <- rr - ss/2
ymb <- ss/2
## arrow locs
x0t <- seq(ss, by=2*ss, len=3)
x1t <- x0t + ss
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row: state
symbols(x=xm[c(1,3,5,7)], y=rep(ymt,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#488fdf",
        inches=FALSE, add=TRUE, lwd=3)
text("Truth", x=-ss, y=ymt, adj=c(0,0.5), xpd=NA,
     cex=2, col="#488fdf")
arrows(x0=x0t,x1=x1t,y0=ymt, col="#488fdf", lwd=3, length=0.12)
## Time or space
arrows(x0=ss/2, x1=cc-ss/2, y0=-ss/3+ss*2,
       length=0.12, lwd=3, xpd=NA)
text("Time", x=cc/2, y=-ss/2+ss*2, xpd=NA, pos=1, cex=2)
```


## Part 1: State model

States of nature might be:

 * animal's location
 
 * density of organisms
 
 * reproductive state
 

## Changes in the state of nature

We can think of changes in the state of nature being driven by a combination of

 * intrinsic (eg, fecundity), and

 * extrinsic factors (eg, temperature)


## Process errors

Some of the extrinsic drivers may be unknown

In time series modeling, we often call these unknown extrinsic factors _process errors_


## Observation (sampling) errors

Consider for a moment the very act of collecting data

Information gathered depends on many factors

* Environmental conditions (eg, cloud cover)

* Behavior (eg, threat avoidance)

* Demographics (age, sex, maturity)

* Sampling design/coverage

* Observer skill


## &nbsp; {data-background=images/sea_lions_aerial.jpg data-background-size=95%}


## Part 2: Observation model

__Data = Truth + Error__
 
```{r obs_diag, fig.height=4, fig.width=8, out.height="100%", out.width="100%", fig.align='center'}
par(mai=c(0.8,0.8,0,0), omi=rep(0,4))
## arrow locs
x0t <- seq(ss, by=2*ss, len=3)
x1t <- x0t + ss
y0b <- rr - ss
y1b <- ss
## empty plot space
plot(c(0,cc), c(0,rr), type="n", xlab="", ylab="",
     xaxt="n", yaxt="n", bty="n")
## top row: state
symbols(x=xm[c(1,3,5,7)], y=rep(ymt,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#488fdf",
        inches=FALSE, add=TRUE, lwd=3)
text("Truth", x=-ss, y=ymt, adj=c(0,0.5), xpd=NA,
     cex=2, col="#488fdf")
## arrows
arrows(x0=x0t,x1=x1t,y0=ymt, col="#488fdf", lwd=3, length=0.12)
## bottom row: obs
symbols(x=xm[c(1,3,5,7)], y=rep(ss/2,4), circles=rep(ss/2,4),
        lty="solid",  fg=NA, bg="#844870",
        inches=FALSE, add=TRUE, lwd=3)
text("Data", x=-ss, y=ss/2, adj=c(0,0.5), xpd=NA,
     cex=2, col="#844870")
## arrows
arrows(x0=xm[c(1,3,5,7)], y0=y0b, y1=y1b,
       col="#c10101", lwd=3, length=0.12)
## Time or space
arrows(x0=ss/2, x1=cc-ss/2, y0=-ss/3,
       length=0.12, lwd=3, xpd=NA)
text("Time", x=cc/2, y=-ss/2, xpd=NA, pos=1, cex=2)
```


## State-space model | General form

State model (autoregressive)

$x_t = f(x_{t-1}, \text{process error})$

<br>

Observation model

$y_t = g(x_t, \text{observation error})$


## State-space model | Random walk

State model

$x_t = x_{t-1} + w_t$ with $w_t \sim \text{N}(0,q)$

<br>

Observation model

$y_t = x_t + v_t$ with $v_t \sim \text{N}(0,r)$


## State-space model | Biased random walk

State model

$x_t = x_{t-1} + u + w_t$ with $w_t \sim \text{N}(0,q)$

<br>

Observation model

$y_t = x_t + v_t$ with $v_t \sim \text{N}(0,r)$


## State-space model | Mean-reverting with $|b| < 1$

State model

$x_t = b x_{t-1} + w_t$ with $w_t \sim \text{N}(0,q)$

<br>

Observation model

$y_t = x_t + v_t$ with $v_t \sim \text{N}(0,r)$


## Consider these data

```{r ex_bias_rw}
TT <- 30

uu <- -0.2
  
ww <- xx <- rnorm(TT)
for(t in 2:TT) {
  xx[t] <- xx[t-1] + uu + ww[t]
}

ee <- rnorm(TT)
yy <- xx + ee

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, ylim = range(xx,yy),
        lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870",
        las = 1, ylab = expression(italic(y[t])))
```


## Two model options

Linear regression 

$y_t = \alpha + \beta t + e_t$

<br>

Biased random walk

$x_t = x_{t-1} + u + e_t$


## Model options | Linear regression

State model

$x_t = \alpha + \beta t$

<br>

Observation model

$y_t = x_t + v_t$ with $v_t \sim \text{N}(0,r)$

## Linear regression fitted values

```{r linear_regr}
theta <- coef(lm(yy ~ seq(TT)))
y_obs <- theta[1] + theta[2] * seq(TT)

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, lwd = 2, type = "o", pch = 16, ylim = range(yy, y_obs),
        cex = 1.5, col = "#844870",
        las = 1, ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(y_obs, lwd = 2, type = "o", pch = 16, col = "#488fdf")
```

<!-- All of the variance in $y_t$ is due to observation error -->


## Observation errors

```{r}
theta <- coef(lm(yy ~ seq(TT)))
y_obs <- theta[1] + theta[2] * seq(TT)

par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, lwd = 2, type = "o", pch = 16, ylim = range(yy, y_obs),  
        cex = 1.5, col = "#844870",
        las = 1, ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(y_obs, lwd = 2, type = "o", pch = 16, col = "#488fdf")
segments(seq(TT), y0 = y_obs, y1 = yy, col = "red")
```

<!-- Errors $v_t = y_t - x_t$ -->


## Model options | Biased random walk

State model

$x_t = x_{t-1} + u + w_t$ with $w_t \sim \text{N}(0,q)$

<br>

Observation model

$y_t = x_t$


## Examples of biased random walks

```{r ex_biased_rw, fig.dim = c(8,4.5)}
x1 <- cumsum(rnorm(30, 0.1, 0.3)) + 10

x2 <- cumsum(rnorm(30, -0.1, 0.3)) + 10

clr1 <- c("#f7fbff",
          "#deebf7",
          "#c6dbef",
          "#9ecae1",
          "#6baed6",
          "#4292c6",
          "#2171b5",
          "#08519c",
          "#08306b")

clr2 <- c("#fff5f0",
          "#fee0d2",
          "#fcbba1",
          "#fc9272",
          "#fb6a4a",
          "#ef3b2c",
          "#cb181d",
          "#a50f15",
          "#67000d")

par(mfrow = c(1,2), mai = c(1, 1, 0.75, 0))
plot.ts(x1, las = 1, col = "dodgerblue", lwd = 2, ylim = c(8, 16),
        ylab = expression(italic(x[t])), main = "")
mtext("Positive bias", side = 3, line = 1, cex = 1.5)
for(i in 9:3) {
  lines(cumsum(rnorm(30, 0.1, 0.3)) + 10,
        col = clr1[i], lwd = 2)
}

plot.ts(x2, las = 1, col = "indianred", lwd = 2,  ylim = c(4, 12),
        ylab = expression(italic(x[t])), main = "")
mtext("Negative bias", side = 3, line = 1, cex = 1.5)
for(i in 9:3) {
  lines(cumsum(rnorm(30, -0.1, 0.3)) + 10,
        col = clr2[i], lwd = 2)
}
```


## Biased random walk

```{r brw_states_obs_0}
xw <- rbind(cbind(seq(TT), xx), cbind(seq(TT)+0.5, xx+uu))
xw <- xw[order(xw[,1]),]
xw[,1] <- c(1, rep(seq(2, TT), ea = 2), TT)
  
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
## blank slate
plot.ts(yy, ylim = range(xx,yy), type = "n",
        las = 1, ylab = expression(italic(x[t])))
lines(xw[,1], xw[,2], lwd = 2, col = "gray")
lines(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf")
```


## Biased random walk fit to earlier data

```{r biased_rw_fit}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870",
        las = 1, ylab = expression(italic(y[t])))
lines(yy, lwd = 2, type = "o", pch = 16, col = "#488fdf")
```

<!-- All of the variance in $y_t$ is due to process error -->


## Process errors

```{r biased_rw_errors}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870",
        las = 1, ylab = expression(italic(y[t])))
lines(yy, lwd = 2, type = "o", pch = 16, col = "#488fdf")
segments(seq(2,TT), y0 = yy[-TT], y1 = yy[-1], col = "red")
```

<!-- Errors: $w_t = y_t - y_{t-1}$ -->


## Model options | Biased random walk with observation error

State model

$x_t = x_{t-1} + u + w_t$ with $w_t \sim \text{N}(0,q)$

<br>

Observation model

$y_t = x_t + v_t$ with $v_t \sim \text{N}(0,r)$


## Biased random walk

```{r brw_states_obs}
xw <- rbind(cbind(seq(TT), xx), cbind(seq(TT)+0.5, xx+uu))
xw <- xw[order(xw[,1]),]
xw[,1] <- c(1, rep(seq(2, TT), ea = 2), TT)
  
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
## blank slate
plot.ts(yy, ylim = range(xx,yy), type = "n",
        las = 1, ylab = expression(italic(x[t])))
lines(xw[,1], xw[,2], lwd = 2, col = "gray")
lines(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf")
```


## Biased random walk

```{r brw_states_obs_2}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
## blank slate
plot.ts(yy, ylim = range(xx,yy), type = "n",
        las = 1, ylab = expression(italic(x[t])))
## lines(xw[,1], xw[,2], type = "o", pch = 16, lwd = 2, col = "gray")
lines(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf")
```


## Biased random walk with obs error

```{r brw_states_obs_errors}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
plot.ts(yy, type = "n",
        las = 1, ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf")
lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870")
segments(seq(TT), y0 = xx, y1 = yy, col = "red")
```


## Biased random walk with obs error

```{r brw_states_obs_errors_2}
par(mai = c(0.9,0.9,0.1,0.1), omi = c(0,0,0,0))
## blank slate
plot.ts(yy, ylim = range(xx,yy), type = "n",
        las = 1, ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(xw[,1], xw[,2], lwd = 2, col = "gray")
lines(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf")
segments(seq(TT), y0 = xx, y1 = yy, col = "red")
lines(yy, lwd = 2, type = "o", pch = 16, col = "#844870", cex = 1.5)
```

The variance in $y_t$ is due to both process & observation error


## Separating process & obs errors

How is it possible to separate out both the process and observation errors?

They have different temporal patterns


## Separating process & obs errors

```{r ex_proc_obs_errors, fig.align="center"}
xp <- matrix(NA, TT, 3)
xp[1,] <- c(0,0,0)

for(t in 2:TT) {
  xp[t,] <- xp[t-1,] + uu + rnorm(3)
}

yp <- xp[,1] + matrix(rnorm(TT*3), TT, 3)

par(mfcol = c(2,3), mai = c(0.6,0.6,0.1,0), omi = c(0,0,0.2,0))
for(i in 1:3) {
  plot.ts(xp[,i], xlab = "", ylab = expression(italic(x[t])))
  if(i == 2) {
    mtext(side = 3, " Different realizations of same process",
          line = 0.5, xpd = NA)
    }
  plot.ts(yp[,i], ylab = expression(italic(y[t])))
  if(i == 2) {
    mtext(side = 3, " Different observations of same process",
          line = 0.5, xpd = NA)
    }
}
```


# Population growth | Density independent


## Density-independent popn growth | Exponential growth/decline in continuous time

$$
N(t) = N_0 ~ \underbrace{\exp(u)}_{\substack{\text{rate of} \\ \text{change}}} \underbrace{\exp(w t)}_{\substack{\text{stochastic} \\ \text{environment}}}
$$


## Density-independent popn growth | In discrete time, with a time step of 1 year

$$
N_t = N_{t-1} \exp(u + w_t)
$$


## Density-independent popn growth | In discrete time, with a time step of 1 year

$$
N_t = N_{t-1} \exp(u + w_t)
$$

<br>

Taking the log of both sides and substituting $x_t$ for $\log(N_t)$

$$
\log (N_t) = \log (N_{t-1}) + u + w_t \\
\Downarrow \\
x_t = x_{t-1} + u + w_t
$$


## Density-independent popn growth | Biased random walk

If we assume that the errors are white noise

$$
w_t \sim \text{N}(0, q)
$$

then our model of populations dynamics

$$
x_t = x_{t-1} + u + w_t
$$

is a biased random walk


## Simulate a biased random walk

```{r rw_sim, echo = TRUE, eval = TRUE}
## number of time steps
TT <- 40
## bias term
uu <- 0.2
## time series of process errors with var = 1
ww <- rnorm(TT, 0, sqrt(1))
## initialize state & set x0 = w0
xx <- ww
## loop over time steps
for(t in 2:TT) {
  xx[t] <- uu + xx[t-1] + ww[t]
}
```


## Simulate a biased random walk

Add some observation error

```{r rw_obs_error, echo=TRUE}
## obs errors with var = 0.3
vv <- rnorm(TT, 0, sqrt(0.3))
## obs data
yy <- xx + vv
```


<!-- ## Plot log-population size over time -->

<!-- Plot the state and observation -->

<!-- ```{r rw_plot_code, echo=TRUE, eval=FALSE} -->
<!-- plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf", -->
<!--         las = 1, ylim = c(min(xx,yy), max(xx,yy)), -->
<!--         ylab = expression(italic(x[t])~~or~~italic(y[t]))) -->
<!-- lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870") -->
<!-- ``` -->


## Log-population size over time

```{r rw_plot_out, align.fig="center"}
par(mai = c(0.8,0.8,0,0), omi = c(0,0,0,0))
plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylim = c(min(xx,yy), max(xx,yy)),
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870")
```


# Population growth | Density dependent


## Density-dependent population growth | Discrete-time stochastic Gompertz model

$$
N_t = N_{t-1} ~ \underbrace{\exp(u ~ + ~ (b - 1) \log(N_{t-1}))}_{\substack{\text{rate of} \\ \text{change}}} \underbrace{\exp(w _t)}_{\substack{\text{stochastic} \\ \text{environment}}}
$$


## Density-dependent population growth | Discrete-time stochastic Gompertz model

$$
N_t = N_{t-1} ~ \exp(u ~ + ~ (b - 1) \log(N_{t-1})) \exp(w _t)
$$

<br>

Taking the log of both sides and substituting $x_t$ for $\log(N_t)$

$$
\begin{align}
\log(N_t) & = \log(N_{t-1}) + u + (b - 1) \log(N_{t-1}) + w_t \\
& \Downarrow \\
x_t &= x_{t-1} + u + (b - 1) x_{t-1} + w_t \\
    &= x_{t-1} + u + b x_{t-1} - x_{t-1} + w_t \\
    &= u + b x_{t-1} + w_t
\end{align}
$$


## Density-dependent population growth | Discrete-time stochastic Gompertz model

We generally assume the process errors are Gaussian

$$
x_t = u + b x_{t-1} + w_t \\
~ \\
w_t \sim \text{N}(0, q)
$$


## Stochastic Gompertz model | A note on the autoregressive parameter $b$

$$
x_t = u + b x_{t-1} + w_t
$$

<br>

For populations under _negative_ effects of density dependence, we require 

$$
-1 < b < 1
$$


## Stochastic Gompertz model | What is the expectation of $x_t$?

$$
x_t = u + b x_{t-1} + w_t \\
w_t \sim \text{N}(0,q)
$$


## Stochastic Gompertz model | What is the expectation of $x_t$?

$$
x_t = u + b x_{t-1} + w_t \\
w_t \sim \text{N}(0,q)
$$

<br>

$$
\text{E}(x_t) = \frac{u}{1 - b}
$$

Hint: We need to think about this over very long (infinite) time scales


## Stochastic Gompertz model | What is the expectation of $x_t$?

Begin by taking the limit as $t \to \infty$

$$
\lim\limits_{t \to \infty} (x_t = u + b x_{t-1} + w_t) \\
\Downarrow \\
\begin{align}
x_\infty &= u + b x_\infty + w_t \\
x_\infty -b x_\infty &= u + w_t \\
(1 - b)x_\infty &= u + w_t \\
x_\infty &= \frac{(u + w_t)}{(1 - b)} 
\end{align}
$$


## Stochastic Gompertz model | What is the expectation of $x_t$?

And then recall that the expectation (mean) of $w_t$ is 0

$$
\begin{align}
\text{E} (x_\infty) &= \text{E} \left( \frac{u + w_t}{1 - b} \right)\\
\text{E} (x_\infty) &= \text{E} \left( \frac{u}{1 - b} \right) + \text{E} \left( \frac{w_t}{1 - b} \right) \\
\text{E} (x_\infty) &= \frac{u}{1 - b} + \frac{0}{1 - b} \\
\text{E} (x_\infty) &= \frac{u}{1 - b}
\end{align}
$$


## Stochastic Gompertz model | Estimating parameters

$$
x_t = u + b x_{t-1} + w_t
$$

It turns out that $u$ and $b$ are confounded and create a likelihood surface with a strong ridge

This means we need a _long_ time series or mulitple observations of the process 

In practice, we will de-mean our data and instead use a familiar AR(1) model

$$
x_t = b x_{t-1} + w_t
$$


## Stochastic Gompertz model

When our population censuses contain observation or sampling errors, we can use a state-space version

$$
x_t = b x_{t-1} + w_t \\
y_t = x_t + v_t
$$


## Simulate a Gompertz model

Simulate the AR(1) state model

```{r gompertz_sim, echo = TRUE, eval= TRUE}
## number of time steps
TT <- 40
## strength of density-dependence (0 < b < 1)
bb <- 0.5
## time series of process errors with SD = 1
ww <- rnorm(TT, 0, 0.5)
## initialize state & set x0 = w0
xx <- ww
## loop over time steps
for(t in 2:TT) {
  xx[t] <- bb * xx[t-1] + ww[t]
}
```


## Simulate a Gompertz model

Add some observation error

```{r gomp_obs_error, echo = TRUE, eval= TRUE}
## obs errors with SD = 0.5
vv <- rnorm(TT, 0, 0.5)
## obs data
yy <- xx + vv
```


<!-- ## Plot log-population size over time -->

<!-- Plot the state and observation -->

<!-- ```{r, echo=TRUE, eval=FALSE} -->
<!-- plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf", -->
<!--         las = 1, ylim = c(min(xx,yy), max(xx,yy)), -->
<!--         ylab = expression(italic(x[t])~~or~~italic(y[t]))) -->
<!-- lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870") -->
<!-- ``` -->


## Log-population size over time

```{r gomp_plot, align.fig="center"}
par(mai = c(0.8,0.8,0,0), omi = c(0,0,0,0))
plot.ts(xx, lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylim = c(min(xx,yy), max(xx,yy)),
        ylab = expression(italic(x[t])~~or~~italic(y[t])))
lines(yy, lwd = 2, type = "o", pch = 16, cex = 1.5, col = "#844870")
```


# Covariates in state-space models


## Covariates in state-space models

We can include covariates (explanatory variables) as well

<br>

Covariates in the state model _affect the underlying process_

$$
x_t = b x_{t-1} + C c_t + w_t
$$


## Covariates in state-space models

We can include covariates (explanatory variables) as well

<br>

Covariates in the state model _affect the underlying process_

$$
x_t = b x_{t-1} + C c_t + w_t
$$

<br>

Covariates in the observation model are _offsets to the underlying process_

$$
y_t = x_t + D d_t + v_t
$$


## Covariates in state-space models | Example with a discrete sine wave as a covariate

State model

$$x_t = 0.7 x_{t-1} + 2 \sin_t + w_t \\ y_t = x_t + v_t$$

<br>

Observation model

$$x_t = 0.7 x_{t-1} + w_t \\ y_t = x_t + 2 \sin_t + v_t$$


## Covariates in the state model

```{r states_covars}
ww <- xx <- xy <- rnorm(TT)

bb <- 0.7

CC <- DD <- 2
cc <- dd <- sin(2*pi*seq(TT)/12)
  
for(t in 2:TT) {
  xx[t] <- bb * xx[t-1] + CC * cc[t] + ww[t]
  xy[t] <- bb * xy[t-1] + ww[t]
}

ee <- rnorm(TT, 0, 1)
yy <- xx + ee

par(mai = c(0.9,0.9,0.5,0.1), omi = c(0,0,0,0))
plot.ts(xx, ylim = range(xx,yy),
        lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylab = expression(paste(italic(x[t]), " or ", italic(y[t]))))
lines(yy, lwd = 2, type = "o", pch = 16, col = "#844870",)
# mtext(side = 3,
#       expression(italic(x[t])==italic(bx[t-1])~+~italic(Cc[t])~+~italic(w[t])),
#       line = 0.5, adj = 0)
```

## Covariates in the observation model

```{r obs_covars}
# ww <- xx <- xy <- rnorm(TT)
# 
# bb <- 0.7
# 
# CC <- DD <- 2
# cc <- dd <- sin(2*pi*seq(TT)/12)
#   
# for(t in 2:TT) {
#   xx[t] <- bb * xx[t-1] + CC * cc[t] + ww[t]
#   xy[t] <- bb * xy[t-1] + ww[t]
# }
# 
# ee <- rnorm(TT)
yy <- xy + DD * dd + ee

par(mai = c(0.9,0.9,0.5,0.1), omi = c(0,0,0,0))
plot.ts(xy, ylim = range(xy,yy),
        lwd = 2, type = "o", pch = 16, col = "#488fdf",
        las = 1, ylab = expression(paste(italic(x[t]), " or ", italic(y[t]))))
lines(yy, lwd = 2, type = "o", pch = 16, col = "#844870",)
# mtext(side = 3,
#       expression(italic(y[t])==italic(x[t])~+~italic(Dd[t])~+~italic(v[t])),
#       line = 0.5, adj = 0)
```

