---
title: "Getting help with errors in R"
subtitle: "Answer Key for Assignment #3"
author: "FISH 549 - Best Practices in Environmental Data Science"
date: "5 February 2023"
output:
  html_document:
    theme: spacelab
    highlight: textmate
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<br>

# Background

Your assignment was to create a reprex for use in a quest for help. It turns out that the problem was that the code was improperly indexing the values upon which we had hoped to compute the mean.

## Setup

Let's walk through the code to find out where our problem lies. Here's the script `01_summary_statistics.R`

```{r problem_script, eval = FALSE}
## this script loads the data and calculates some summary statistics

## load libraries
library("here")

## set location of the data directory 
data_dir <- here("data")

## load data file
pisaster_data <- readRDS(file.path(data_dir, "pisaster_data.Rds"))

## peek at the data
head(pisaster_data)

## calculate mean counts across all years, sites, and plots
mean_count <- mean(pisaster_data$count)
```

```{r problem_script_load, echo = FALSE}
## this script loads the data and calculates some summary statistics

## load libraries
library("here")

## set location of the data directory 
data_dir <- here("homework", "week_03")

## load data file
pisaster_data <- readRDS(file.path(data_dir, "pisaster_data.Rds"))

## peek at the data
head(pisaster_data)

## calculate mean counts across all years, sites, and plots
mean_count <- mean(pisaster_data$count)
```

The warning message suggests that the object we're passing to `mean()` is not numeric, which seems odd at first glance because `count` does appear to be numeric in that its values are not surrounded by quotes as with `site`.

<br>

# Create a reprex

Here's an example of what my reprex would look like. The key elements here are:

* the question/problem is stated clearly and succinctly

* a short code snippet with *only the info necessary to address the error*

* the data provided as a `structure`

* all of the information is contained here (question/problem, code, data)

```{r eval = FALSE}
I'm trying to compute the mean of a column in a data frame, but I get the following error:

Warning in mean.default(ex_data$count): argument is not numeric or
logical: returning NA

Here is a shortened version of the data structure:

## example data
ex_data <- structure(list(2019, 2019, 2019, 2019, 2019, 2019, 2019, 2019, 
    2019, 2019, "a", "a", "a", "a", "a", "b", "b", "b", "b", 
    "b", 1L, 2L, 3L, 4L, 5L, 1L, 2L, 3L, 4L, 5L, 
    6L, 10L, 13L, 9L, 9L, 11L, 7L, 10L, 12L, 
    8L), .Dim = c(10L, 4L), .Dimnames = list(
    NULL, c("year", "site", "plot", "count")))

Here is the problematic line of code:

## calculate the mean count across all times/sites/plots
mean_count <- mean(ex_data$count)

Here is my `sessionInfo()`

R version 4.0.3 (2020-10-10)
Platform: x86_64-apple-darwin17.0 (64-bit)
Running under: macOS Mojave 10.14.6

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib
LAPACK: /Library/Frameworks/R.framework/Versions/4.0/Resources/lib/libRlapack.dylib

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base  
```

<br>

# Solution to the problem

The warning itself suggests that there is something wrong with the object we're passing to `mean()` (i.e., `argument is not numeric or logical`), so the first thing to do would be to check on the class of the object.

```{r}
## get the example data
ex_data <- structure(list(2019, 2019, 2019, 2019, 2019, 2019, 2019, 2019, 
    2019, 2019, "a", "a", "a", "a", "a", "b", "b", "b", "b", 
    "b", 1L, 2L, 3L, 4L, 5L, 1L, 2L, 3L, 4L, 5L, 
    6L, 10L, 13L, 9L, 9L, 11L, 7L, 10L, 12L, 
    8L), .Dim = c(10L, 4L), .Dimnames = list(
    NULL, c("year", "site", "plot", "count")))

## peek at the data
head(ex_data)

## what's the class of the object?
class(ex_data$count)
```

That's odd. We can clearly see there are integers in the `count` column (denoted by the `L` in the `structure(...)` command). What's the class of the larger object `ex_data` that `count` is a part of?

```{r}
## what's the class of the larger object?
class(ex_data)
```

OK, `ex_data` is not actually a `data.frame`, but rather a `matrix` (or equivalently a 2D `array`). That means we can't index the columns with a dollar sign (`$`) and instead must use numeric or character row/column indexing (e.g., `ex_data[, 4]` or `ex_data[, "count"]`).

Let's try to calculate the mean that way.

```{r}
## Trying a different indexing
mean(ex_data[, "count"])
```

Hmm, still not working. Apparently the object is still not `numeric`. Let's examine the class of the object with the proper indexing.

```{r}
## inspect class
class(ex_data[, "count"])
```

This is getting weird. The column of numbers in `count` is apparently a `list`. Let's take a look.

```{r}
## inspect the object
ex_data[, "count"]
```

Sure enough. Each of the numbers in the `count` column is an element of a list, but we can't access the list with a standard `$` operator. How does that work?

The trick here (and often elsewhere) is to examine the structure of the example data with `str()`.

```{r}
str(ex_data)
```

This clearly shows that `ex_data` is a list containing 40 objects, but unlike a standard list whose `dim()` should be `NULL`, this object has dimensions of 10 rows by 4 columns (i.e., `attr(*, "dim")= int [1:2] 10 4`). That's why `class(ex_data) == c("matrix", "array")` in the code above. The solution here is to use `unlist()` on `count()` before passing it to `mean()`.

```{r}
## load `magrittr` to get the pipe operator
library(magrittr)

## SOLUTION: calculate mean counts across all years, sites, and plots
mean_count <- ex_data[,"count"] %>%
  unlist() %>%
  mean()
mean_count
```

<br>

# Endnote

There are 3 more hints that `ex_data` is not a `data.frame`. First, the character values in `site` have quotes around them instead of simply being displayed as normal text. For example, here is a comparison of `ex_data` before and after converting it to a `data.frame`:

```{r}
## original
ex_data

## as a data frame
as.data.frame(ex_data)
```

Second, the row names in a matrix appear as standard row notation (`[#,]`) rather than a simple integer (`#`) as in a data frame. Third, all of the values in the columns are left-justified in the original matrix form of `ex_data`, but right-justified in the data frame.

The data structure I used here is rather uncommon and something that very few people would ever use. That is, I created a `matrix` wherein each element was itself a `list` with only one value in it. In so doing, I was able to combine integer/numeric and character values into the same matrix, which otherwise would not work. For example, the following line of code will convert the numeric `1` and `2` to the characters `"1"` and `"2"`.

```{r}
## create matrix
(mm <- matrix(c(1, 2, "a", "b"), 2, 2))

## check class of rows
mm %>% apply(1, class)

## check class of cols
mm %>% apply(2, class)
```

<br><br><br><br><br>

