---
title: "Code reviews"
subtitle: "Best Practices in Environmental Data Science<br>FISH 549"
date: "6 March 2023"
output:
  ioslides_presentation:
    css: lecture_slides.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is a code review?

Inspecting someone's code with an eye towards improvement


## What is a code review?

Inspecting someone's code with an eye towards improvement

It's not unlike peer review of scientific publications


## Actors

**Author**: writes the code and sends it for review

<br>

**Reviewer**: reads the code and makes comments/suggestions


## The ask

Author and reviewer should be clear on roles & expectations

> - Be specific ("Please look over my changes to function `f()` on lines 11-20")

> - Be aware of subject expertise

> - Be thoughtful of time ("This should only take a few minutes")


## The response

Be honest and direct

> - address the intent of the review

> - don't redo the analysis


## Areas covered

* correctness of the code (are there bugs or errors?)

> - test coverage (do unit tests span the full range of options?)

> - functionality changes (does change to `f()` affect `g()`?)

> - coding guides (does it follow best practices?)


## Tone

"Be tough on the code, but easy on the author"

> - be empathetic, kind & positive

> - consider both intent and impact

> - ask open-ended questions


## Avoid "you"

What you said:

* "you misspelled `easy`"


## Avoid "you"

What you said:

* "you misspelled `easy`"

What they heard:

* Option 1: "Oops--honest typo"

> - Option 2: "You are dumb"


## Requests vs commands

Consider these 2 statements:

1) "Change `A` to `a`"

2) "Perhaps we could change uppercase parameter names to lowercase?"

<br>

\#1 is *combative* whereas \#2 is *cooperative*


## Principles vs opinions

Attach reasons to requests

> - "We should consider lowercase letters for parameter names to distinguish them from variable names that are generally written in uppercase"


## Bring solutions

Receiving feedback can be tough, so offer concrete solutions to soften the blow

> - "I found this code block hard to understand. What if we used pipes `%>%` instead of nested functions?"


## High to low

Begin reviews with high-level comments

* "This might be more clear to others if we split `f()` into `g()` and `h()`"


## High to low

Begin reviews with high-level comments

* "This might be more clear to others if we split `f()` into `g()` and `h()`"

and then focus on details

* "We could change line 11 to instead be `b <- a + 1`"


## Keep it in scope

You're asked to review function `f()`

> - only consider `g()` if it affects `f()`

> - avoid "`f()` looks great but I found a bug in `h()`"


## Limit comments

Imagine `easy` is misspelled everywhere

> - Option 1: flag *every* occurrence with "change `eesy` to `easy`"

> - Option 2: "Here (and elsewhere) we should change `eesy` to `easy`"


## Nitpicks | Unimportant comments

* often involve adherence to a style guide

> - alphabetical or hierarchical structuring of declarations

> - placement of comments, brackets, function args


## Nitpicks {.smaller}

```{r eval = FALSE, echo = TRUE}
a <- 1
b <- 2
a + b
c <- 3
(c - a) / b
```

becomes

```{r eval = FALSE, echo = TRUE}
## parameter definitions
a <- 1
b <- 2
c <- 3

## step 1: addition
a + b

## step 2: division
(c - a) / b
```


## A note of caution

<br>

<div class="centered" style="font-size:1.5em">
**Nitpicks can be microaggressions**
</div>


## Humility

*[P]rogrammers are very unlikable people... In aviation, for example, people who greatly overestimate their level of skill are all dead.*

-- Philip Greenspun


## The sweetest of all sounds is praise

By design, reviews target faults

Litter your review with the things you like

> - "I'd never seen this trick before!"

> - "This is a much more elegant solution than I would have used."

> - "I really like how you broke this problem down."


## Communication

Most code reviews are done asynchronously (eg, pull request on GitHub)

> - leave meaningful comments & ask judicious questions

> - *N*<sub>comments</sub> $\propto$ level of misunderstanding

> - at some point, a spoken conversation is best


## New people

Newcomers can be overwhelmed with styles, culture, people

> - good reviews apply the same metrics/qualities to **everyone**

> - try to place people in a position to succeed


## Support

Organizations or labs should set clear expectations and guidelines

> - What are people's roles?

> - What are expectations for people's time?

> - Who has final say? Individual or Team?


## Innovation

Improvements should not be limited to code, but to processes as well

> - "lots of people are confused by this"

> - "we keep making the same mistake"


# Reviews in practice


## How?

Embrace collaboration platforms like GitHub

> - Issues

> - Pull requests

> - Gists


## Simple requests

Options include:

* code snippet in an email

* post to Slack


## Formal requests

Recall our discussion of *minimal working examples*

> - general cry for help

> - specific ask


## rOpenSci

"fosters a culture that values **open** and **reproducible** research using **shared** data and reusable software"


## rOpenSci

Uses GitHub's infrastructure with a focus on issues for package reviews


## {data-background=img/ros_readme.png data-background-size=80% data-background-position="50% 40%"}


## Reviewer {data-background=img/review_template.png data-background-size=60% data-background-position="75% 50%"}


## Editor {data-background=img/editor_template.png data-background-size=75% data-background-position="50% 50%"}


## Summary

Be kind

Be thorough

Be honest

Be supportive

Be timely


