---
title: "Forking a repo and submitting a pull request"
subtitle: "Assignment #2"
author: "FISH 549 - Best Practices in Environmental Data Science"
date: "20 January 2023"
output:
  html_document:
    theme: spacelab
    highlight: textmate
    toc: false
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>

# Forking workflows

This assignment will allow you to demonstrate your ability to "fork" a repository on GitHub, clone it to RStudio, make some local changes, push them to GitHub, and submit a "pull request". If you need a reminder as to how this process works, please refer to the notes [here](https://fish549.github.io/website/lectures/week_03/lec_08_workflows_ex.html#Forking_workflow).

***

# Task 1

Navigate to the repo https://github.com/FISH549/assignment-2 and fork it to your own GitHub account.

***

# Task 2

Clone your forked copy of the repo to your own computer by creating a new project in RStudio.

***

# Task 3

Create a new branch in your local repo called `develop`.

***

# Task 4

In the `develop` branch, create a new R script with the following information:

```{r eval = FALSE}
## A simple function for adding 2 numbers
add <- function(a, b) {
  return(a + b)
}
```

Save this script as `add.R` and then commit it to your local repo. When you are finished, push this file to your remote repo on GitHub.

***

# Task 5

Navigate to your remote repo on GitHub to review your new file and submit a pull request (PR). Give your PR an informative title and leave a short description of what your new file does. When you're finished, submit the PR for Mark's review.

